### 2015-01-20 doc/thesis/typedlua.tex
Minor improvements to the text about types and subtyping
Improving the text about optional type annotations, function types, and union types

Typed Lua is an optional type system for Lua, and its main goal is to
allow programmers to combine static and dynamic typing in Lua.
In practice, Typed Lua extends the syntax of Lua to introduce
optional type annotations.
In addition, Typed Lua performs local type inference \citep{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

An important feature of Typed Lua is its backwards compatibility with Lua,
that is, unannotated Typed Lua code is valid Lua code.
For this reason, the syntactic extensions introduced by Typed Lua
include new keywords that are not reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
in extended BNF.

Another important feature of Typed Lua is its interface between
statically and dynamically typed modules.
This means that the programmer can enjoy some of the benefits of
static types even without converting existing Lua modules to Typed Lua,
as a dynamically typed module can export a statically typed interface,
and statically typed users of the module have their use of the module
checked by the compiler.

We use the consistent-subtyping relation of gradual typing
\citep{siek2007objects,siek2013mutable} to formalize Typed Lua,
though it does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
A good reason for not inserting run-time checks at this moment is that they
can decrease run-time performance \citep{allende2013cis}, and we believe that
a careful evaluation of run-time checks should be done before inserting them
on the type system.

Unlike Dart \citep{dart} and TypeScript \citep{typescript},
we are designing Typed Lua aiming soundness to make it possible to switch
Typed Lua from optional typing to gradual typing in the future,
as a sound type system is a prerequisite to make Typed Lua insert
run-time checks after static type checking.

In this chapter we use some examples of Typed Lua code to show how
they relate to Lua.
This examples shall give an informal overview of our optional type system.
We use typing rules to present the formalization of our optional type system
in Chapter \ref{chap:system}.
It worth mentioning that we used our implementation to produce and
test all the examples that we present along this section.

@@

Lua values can have one of eight tags:
\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
Typed Lua includes types for the first six.
Typed Lua also includes a syntactical extension that programmers can use
to define the types of \emph{userdata}, and uses it to define the type \emph{thread}. 
In this section, we present the Typed Lua types that may appear on annotations.
We explain all Typed Lua types and syntactical extensions along this chapter.

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type} \; | \;
  \textit{recordtype}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{any}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

Figure \ref{fig:types} presents the concrete syntax of the
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include the type void \texttt{()},
vararg types, tuple types, and unions of tuple types.

Typed Lua uses subtyping to order types.
Any first-level type is a subtype of \texttt{value}.
The rest of the subtyping relationship is standard:
union types are supertypes of their parts,
\texttt{boolean}, \texttt{integer}, \texttt{number}, and \texttt{string}
are supertypes of their respective literal types,
\texttt{integer} is subtype of \texttt{number},
function types are related by contravariance on the input
and covariance on the output, table types have width subtyping,
with depth subtyping on \texttt{const} fields, tuple and vararg types
are covariant.

Typed Lua uses consistent-subtyping to check the interaction among the
dynamic type \texttt{any} and other types.
The dynamic type \texttt{any} is a subtype of \texttt{value}, but neither
a supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and can pass any value where a
value of the dynamic type is expected, but the compiler tracks these
operations, and the programmer can choose to be warned about them.

Typed Lua allows optional type annotations in variable and function
declarations.
Whenever it is possible to use local type inference, the compiler
assigns a more specific type to an unannotated declaration.
When it is not possible to use local type inference, the compiler
assigns the type \texttt{any} to an unannotated declaration.
For instance, in function declarations, the compiler does not infer
a more specific type to unannotated input parameters, because,
in this case, global type inference often is undecidable in a
language with subtyping and overloading \citep{wells1999typability}.

In the following example, we use type annotations in a function
declaration, and we do not use type annotations in a declaration of a
local variable:
\begin{verbatim}
    local function factorial (n:integer):integer
      if n == 0 then
        return 1
      else
        return n * factorial(n - 1)
      end
    end
    local x = 5
    print(factorial(x))
\end{verbatim}

Even though we did not add type annotations in the declaration of
the local variable \texttt{x}, the compiler uses local type inference to
assign the type \texttt{integer} to \texttt{x}, and this example compiles
without any errors.
The compiler does not use the literal type \texttt{5},
instead of the base type \texttt{integer}, because this would prevent
programmers from assigning other integer values to the variable
\texttt{x}.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.

Typed Lua allows programmers to combine statically typed code with
dynamically typed code, as we show in the following example:
\begin{verbatim}
    local function absolute (n:integer):integer
      if n < 0 then
        return -n
      else
        return n
      end
    end

@@

The function \texttt{distance} receives two parameters of type \texttt{any}
and returns a value of type \texttt{integer}.
The compiler assigns the dynamic type \texttt{any} to the input
parameters of \texttt{distance} because they do not have type annotations,
and the compiler does not use global type inference, as we mentioned
previously.
Even though we did not annotated the return type of \texttt{distance},
the compiler is able to infer its return type because it is local
and not recursive.

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not an integer number,
because in the semantics of Lua the minus operator can result in a
value that is not an integer number.
In fact, we can overload the minus operator to return a value that is
not even a number.
However, we can call \texttt{absolute} inside \texttt{distance}
because the expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{integer}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not an integer, making the dynamically typed code crash the
statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

@@

Lua has first-class functions, but they have some peculiarities.
First, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
Second, functions can return any number of values,
and this number of returned values may not be statically known.
Third, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment
(or vice-versa), that is, calling a function is like doing a
multiple assignment where the left hand side is the parameter list
and the right hand side is the argument list.

Typed Lua uses second-level types to encode function types,
and to preserve these peculiarities.
We call them second-level because these types do not correspond
to actual Lua values and we cannot use them to type variables or
parameters.
They are just an artifact of the interaction between the type system
and the semantics of Lua.

As we can see in Figure \ref{fig:types}, second-level types consist
of either \emph{tupletype} or \emph{rettype}.
A second-level type is the type void \texttt{()},
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
Union of tuples may appear only on the return type of functions,
and together with \emph{projection types} play an important role
in functions that are overloaded on the return type.
We will explain union types and projection types in more detail
in the next section.

To match the semantics of Lua function calls, the \emph{default} operation
mode of Typed Lua adds a variadic tail to the parts of a function
type whenever the programmer does not specify one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(integer, value*) -> (integer, nil*)},
and the type of \texttt{distance} is \texttt{(any, any, value*) -> (integer, nil*)}.
If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
them, as the type signature lets \texttt{absolute} receive any number of
extra arguments.
If we call \texttt{absolute} with fewer arguments, Typed Lua uses \texttt{nil} to
replace the missing arguments and in the case of \texttt{absolute} generates
an error, as \texttt{nil} is neither a subtype of \texttt{integer} nor consistent
with \texttt{integer}.

Typed Lua also provides a \emph{strict} operation mode, and programmers can
use this mode if they want to catch arity mismatch of function calls.
In the strict mode, Typed Lua does not add a variadic tail to the
parts of a function type unless the programmer explicitly declares one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(integer) -> (integer)},
and the type of \texttt{distance} is \texttt{(any, any) -> (integer)}.
If we call \texttt{absolute} with either extra or missing arguments,
Typed Lua generates errors about arity mismatch.

A variadic type can only appear in the tail position of a tuple,
because Lua takes only the first value of any expression that appears
in an expression list that is not in tail position.
The following example shows the interaction between multiple returns
and expression lists:
\begin{verbatim}
    local function multiple ()
      return 2, "foo"
    end
    local function sum (x:integer, y:integer)
      return x + y
    end
    local x, y, z = multiple(), multiple()
    print(sum(multiple(), multiple()))
\end{verbatim}

In the default mode, the type of \texttt{multiple} is
\texttt{(value*) -> (integer, string, nil*)},
and the type of \texttt{sum} is \texttt{(integer, integer, value*) -> (integer, nil*)}.
In the right hand side of the multiple assignment, only the first
value produced by the first call to multiple gets used,
so the type of the right hand side is
\texttt{(integer, integer, string, nil*)},
and the types assigned to \texttt{x}, \texttt{y}, and \texttt{z}
are \texttt{integer}, \texttt{integer}, and \texttt{string}, respectively.
The same tuple is assigned to the call to \texttt{sum} and it
compiles without errors, as the first two components of the tuple
are consistent with the types of the parameters, and the other
components are consistent with \texttt{value}.

In the strict mode, the type of \texttt{multiple} is \texttt{() -> (integer, string)},
and the type of \texttt{sum} is \texttt{(integer, integer) -> (integer)}.
Like in the default mode, only the first value produced by the
first call to multiple gets used in the right hand side of the
multiple assignment;
and the type of the right hand side is also
\texttt{(integer, integer, string, nil*)},
because the strict mode affects only the type of function declarations
and multiple assignments that appear in function calls.
Unlike in the default mode, the type of the call to \texttt{sum} is
\texttt{(integer, integer, string)}, and does not include the
variadic tail to catch arity mismatch.
Thus, this example compiles with error because the call to \texttt{sum}
passes an unexpected extra argument of type \texttt{string}.

@@

    print(message("Typed Lua"))         -- Typed Lua
    print(message("Typed Lua", "Hi "))  -- Hi Typed Lua
\end{verbatim}

Although the parameter \texttt{greeting} is optional, and has type
\texttt{string|nil}, the concatenation does not generate a type error
because we used the short-circuiting \texttt{or} operator to declare a
new variable \texttt{greeting} that is guaranteed to have type \texttt{string}. 
In Lua, only the values \texttt{nil} and \texttt{false} represent a
false condition, so programmers often use the \texttt{or} operator as a
common idiom to assign a default value to an optional parameter. 
Typed Lua uses a typing rule to encode this idiom: 
if the left hand side of \texttt{or} has type \texttt{t|nil} and
the right hand side has type \texttt{t} then the \texttt{or} expression has
type \texttt{t}.

@@

Typed Lua allows the assignment \texttt{x = x or e} to refine
the type of \texttt{x} from \texttt{t|nil} to \texttt{t}
if it matches the following restrictions: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{x} is local to the current function
and it is not being assigned in another function.
The refinement only affects the type of \texttt{x} in the remainder
of the current scope.
In the case of \texttt{greeting}, the assignment refines its type
from \texttt{string|nil} to \texttt{string}.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the \texttt{type(x) == "string"} predicate that
constrains the type of \texttt{x} from \texttt{t|string} to
\texttt{string} when the predicate is true and \texttt{t} otherwise.
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and usually returns \texttt{nil} as the first value plus an
error message or other data that describes the error
when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:integer, divisor:integer):
          (integer, integer)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) // divisor
        return q, r
      end 
    end
\end{verbatim}

@@

When Typed Lua finds a union of tuples in the right hand side of
a declaration, it stores this union in a special type environment
with a fresh name and assigns \emph{projection types} to the
variables that appear unannotated in the left hand side of the declaration.
In our example, Typed Lua assigns the union of tuples
\texttt{(integer, integer, nil*)|(nil, string, nil*)} to a type variable $X$.
Thus, variable \texttt{q} gets the type $X_{1}$
that is equivalent to the type \texttt{integer|nil},
and variable \texttt{r} gets the type $X_{2}$
that is equivalent to the type \texttt{integer|string}.

>>>>

Typed Lua is an optional type system for Lua, and its main goal is to
provide static type checking for Lua.
To do that, Typed Lua extends the syntax of Lua 5.3 to introduce
optional type annotations, and performs local type inference \citep{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

Another goal of Typed Lua is to be backwards compatible with Lua.
This means that any Lua code is valid Typed Lua code.
To be backwards compatible with Lua, the syntactic extensions introduced
by Typed Lua does not include new reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
in extended BNF.

We use the consistent-subtyping relation of gradual typing
\citep{siek2007objects,siek2013mutable} to formalize Typed Lua,
though it does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
We did not insert run-time checks at this moment because they can decrease
run-time performance \citep{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done before
inserting them on the type system.
However, this evaluation is out of the scope of this work.

Unlike Dart \citep{dart} and TypeScript \citep{typescript},
we are designing Typed Lua aiming soundness to make it possible to
switch Typed Lua from optional typing to gradual typing in the future.
As we mentioned in the previous chapter,
a sound type system is a prerequisite to insert run-time checks after
static type checking, because a sound type system ensures that
statically typed code will not throw type errors during run-time.

In this chapter we use some examples of Typed Lua code to show how
they relate to Lua.
These examples shall give an informal overview of our optional type system.
We will use typing rules to present the formalization of our optional type
system in the next chapter.
All the examples that we present along this chapter run in our Typed Lua compiler.

@@

Lua values can have one of eight tags:
\emph{nil}, \emph{boolean}, \emph{number}, \emph{string},
\emph{function}, \emph{table}, \emph{userdata}, and \emph{thread}.
Typed Lua includes types for the first six.
Typed Lua also includes a syntactical extension that programmers can use
to define the types of \emph{userdata}.
We use this syntactical extension to define the type \emph{thread}. 
In this section we present the Typed Lua types that may appear on annotations.
We explain all Typed Lua types and syntactical extensions along this chapter.

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; \textit{hashtype} \; | \;
  \textit{recordtype}\\
\textit{hashtype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

Figure \ref{fig:types} presents the concrete syntax of
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include the type void \texttt{()},
vararg types, tuple types, and unions of tuple types.

Typed Lua uses subtyping to order types.
Any first-level type is a subtype of \texttt{value}.
Union types are supertypes of their parts.
The base types \texttt{boolean}, \texttt{integer}, \texttt{number},
and \texttt{string} are supertypes of their respective literal types.
The base type \texttt{integer} is subtype of \texttt{number}.
Function types are related by contravariance on the input
and covariance on the output.
Table types have width subtyping, with depth subtyping on
\texttt{const} fields.
Tuple and vararg types are covariant.
Unions of tuple types are also supertypes of their parts.

Typed Lua uses consistent-subtyping to check the interaction among the
dynamic type \texttt{any} and other types.
The dynamic type \texttt{any} is a subtype of \texttt{value}, but it is
neither a supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and we can pass any value where a
value of the dynamic type is expected, but the compiler tracks these
operations, and the programmer can choose to be warned about them.

Typed Lua allows optional type annotations in variable and function
declarations.
We use the following example to illustrate how we can annotate a
function declaration and a variable declaration:
\begin{verbatim}
    local function succ (x:integer):integer
      return x + 1
    end
    local x:integer = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

Typed Lua uses local type inference to assign more specific types to
some unannotated declarations.
More precisely, Typed Lua can infer the type of local variables and
the return type of local functions that are not recursive.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.
For local variables, Typed Lua uses the type of the initialization
expression to assign a more specific type to an unannotated local variable.
For local functions, Typed Lua uses the type of the returned expression
to assign a more specific type to an uannotated return type.
This means that we can rewrite the previous example as follows:
\begin{verbatim}
    local function succ (x:integer)
      return x + 1
    end
    local x = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

In this example, the compiler uses local type inference to assign the
type \texttt{integer} to the local variable \texttt{x} and to the
return type of the local function \texttt{succ}, making this example
compile without any warnings.
Local type inference always use the most general type.
In this example, the compiler does not use the literal type \texttt{7},
instead of the base type \texttt{integer}, because this would generate
a warning when we try to assign other integer value to the variable \texttt{x}.
Still, programmers can use literal types in type annotations if
they need a variable that has a very specific type.
In Section \ref{sec:tables} we will see in more detail that literal
types are essential to type Lua tables.

Typed Lua assigns the type \texttt{any} to the unannotated declarations
that it cannot infer a more specific type.
More precisely, Typed Lua cannot infer more specific types to input
parameters of function declarations and to the return type of
recursive functions, as global type inference is undecidable in a
language with subtyping \citep{wells1999typability}.

We use the following example to illustrate type annotations in the
declaration of a recursive function:
\begin{verbatim}
    local function factorial (n:integer):integer
      if n == 0 then
        return 1
      else
        return n * factorial(n - 1)
      end
    end
\end{verbatim}

This example compiles without any warnings because we annotated
the return type of \texttt{factorial}.

We use the following example to illustrate the omission of type
annotations in the input parameters of a local function,
and also to show that Typed Lua allows programmers to combine
statically typed code with dynamically typed code:
\begin{verbatim}
    local function absolute (n:integer):integer
      if n < 0 then
        return -n
      else
        return n
      end
    end

@@

The function \texttt{distance} receives two parameters of type \texttt{any}
and returns a value of type \texttt{integer}.
The compiler assigns the dynamic type \texttt{any} to the input
parameters of \texttt{distance} because they do not have type annotations
and the compiler does not use global type inference, as we mentioned
previously.
Even though we did not annotate the return type of \texttt{distance},
the compiler is able to infer its return type because it is local
and not recursive.

In this example, Typed Lua cannot guarantee that \texttt{distance} is never
going to call \texttt{absolute} with a parameter that is not an integer,
because in the semantics of Lua the minus operator can result in a
value that is not an integer number.
In fact, we can overload the minus operator to return a value that is
not even a number.
However, we can call \texttt{absolute} inside \texttt{distance}
because the expression \texttt{x - y} has type \texttt{any},
and it is consistent with type \texttt{integer}.
Still, the dynamic type \texttt{any} may be hiding a value of a type
that is not an integer, making the dynamically typed code break the
guarantees provided by the statically typed code.
This is a typical example where run-time checks would ensure safety
between the interaction of dynamically typed and statically typed code.

@@

Lua has first-class functions, but they have some peculiarities.
First, the number of arguments of a function call does
not need to match the arity of the function declaration, as Lua silently
drops extra arguments after evaluating them, or uses \texttt{nil} to
replace missing arguments.
Second, functions can return any number of values,
and this number of returned values may not be statically known.
Third, Lua has multiple assignment, and the semantics
of argument passing is the same of the multiple assignment, that is,
calling a function is like doing a multiple assignment where the
left hand side is the parameter list and the right hand side is the
argument list.

Typed Lua uses second-level types to encode function types
and to preserve these peculiarities.
We call them second-level because these types do not correspond
to actual Lua values and we cannot use them to type variables or
parameters.
Second-level types represent tuple types that can appear in multiple assignment.
Since the semantics of argument passing is the same of multiple assignment,
second-level types also appear in function types.

\begin{figure}[!ht]
\textbf{Function types}\\
\dstart
\begin{align*}
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua function types}
\label{fig:functions}
\end{figure}

As we can see in Figure \ref{fig:functions}, second-level types consist
of either \emph{tupletype} or \emph{rettype}.
A second-level type is either the type void \texttt{()},
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
The type void \texttt{()} represents the type of the empty tuple.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
Union of tuple types appear in the return type of function types to
represent overloading on the return type.
We will explain union types in more detail in the next section.
We can use only one first-level type \texttt{t} in the return type
because it is syntactic sugar to the tuple type \texttt{(t)}.

Typed Lua provides two operation modes: the \emph{default} mode and
the \emph{strict} mode.
In the default mode, the compiler adds a variadic tail to the
type of the parameter list and to the return type whenever the
programmer does not specify one.
The default mode has this behavior to preserve the semantics
of function calls in Lua, that is, it discards extra arguments and
uses \texttt{nil} to replace missing arguments.
In the strict mode, the compiler does not add a variadic tail
to the type of the parameter list and to the return type.
Instead, it preserves the type defined by the programmer to catch
arity mismatch.
We will use the following function to illustrate how these two
operation modes work:
\begin{verbatim}
    local function sum (x:integer, y:integer):integer
      return x + y
    end
\end{verbatim}

In the default mode, the type of \texttt{sum} is
\texttt{(integer, integer, value*) -> (integer, nil*)}.
The compiler adds \texttt{value*} to the type of the parameter list
to discard extra arguments.
For instance, the call \texttt{sum(1, 2, 3)} compiles without any
warnings because Typed Lua uses \texttt{value*} to drop the extra
argument \texttt{3}.

In the strict mode, the type of \texttt{sum} is
\texttt{(integer, integer) -> (integer)}.
The compiler does not add \texttt{value*} to the type of the parameter
list to catch arity mismatch.
For instance, the call \texttt{sum(1, 2, 3)} compiles with a warning
because we are passing an extra argument \texttt{3} to \texttt{sum}.

Even though these operation modes affect function calls, they do not
affect other multiple assignment.
For instance,
\begin{verbatim}
    local x:integer, y:integer = 1, 2, 3
\end{verbatim}
compiles without any warnings in both operation modes.
This means that the left-hand side of this local declaration has the
tuple type \texttt{(integer, integer, value*)}, while the expression
list in the right-hand side of this local declaration has the
tuple type \texttt{(1, 2, 3, nil*)} in both operation modes.

A variadic type can only appear in the tail position of a tuple,
because Lua takes only the first value of any expression that appears
in an expression list that is not in tail position.
We will use the following function to illustrate the interaction
between multiple returns and expression lists:
\begin{verbatim}
    local function m ():(integer, string)
      return 2, "foo"
    end
\end{verbatim}

As an example,
\begin{verbatim}
    local x:integer, y:integer, z:string = m(), m()
\end{verbatim}
compiles without warnings in both operation modes.
This happens because in the right hand side of the multiple
assignment, only the first value produced by the first call
to \texttt{m} gets used, so the type of the right hand side
is \texttt{(integer, integer, string, nil*)}, which is
consistent with the type of the left hand side
\texttt{(integer, integer, string, value*)}.

We can also type variadic functions.
For instance,
\begin{verbatim}
    local function v (...:string):(string*)
      return ...
    end
\end{verbatim}
has type \texttt{(string*) -> (string*)} in both operation modes.
The call \texttt{v()} compiles without any warnings in the default and
strict modes because \texttt{(value*)} and \texttt{()} are consistent
with \texttt{(string*)}.
The call \texttt{v("hello", "world")} compiles without any warnings
in the default and strict modes because \texttt{("hello", "world", nil*)}
and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
The call \texttt{v(...)} compiles withouth any warnings in both
operation modes because the type of the argument list is \texttt{(string*)},
assuming that the vararg expression (\texttt{...}) has type \texttt{string}.

@@

    print(message("Typed Lua"))         --> Typed Lua
    print(message("Typed Lua", "Hi "))  --> Hi Typed Lua
\end{verbatim}

Although the parameter \texttt{greeting} is optional, and has type
\texttt{string|nil}, the concatenation does not generate a warning
because we used the short-circuiting \texttt{or} operator to declare a
new variable \texttt{greeting} that is guaranteed to have type \texttt{string}. 
In Lua, only the values \texttt{nil} and \texttt{false} represent a
false condition, so programmers often use the \texttt{or} operator as a
common idiom to assign a default value to an optional parameter. 
Typed Lua uses the following rule to type this idiom: 
if the left hand side of \texttt{or} has type \texttt{t|nil} and
the right hand side has type \texttt{t} then the \texttt{or} expression has
type \texttt{t}.

@@

Typed Lua allows the assignment \texttt{x = x or e} to refine
the type of \texttt{x} from \texttt{t|nil} to \texttt{t}
if it matches the following rules: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{x} is local to the current function
and it is not being assigned in another function.
The refinement only affects the type of \texttt{x} until the
next assignment to \texttt{x}.
In the case of \texttt{greeting}, the assignment refines its type
from \texttt{string|nil} to \texttt{string}.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the \texttt{type(x) == "string"} predicate that
constrains the type of \texttt{x} from \texttt{string|t} to
\texttt{string} inside the \texttt{if} body and to \texttt{t}
inside the \texttt{else} body. 
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and returns \texttt{nil} plus an error message or other data that
describes the error when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:integer, divisor:integer):
          (integer, integer)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) // divisor
        return q, r
      end 
    end
\end{verbatim}

@@

When Typed Lua finds a union of tuples in the right hand side of
a declaration, it assigns \emph{projection types} to the
variables that appear unannotated in the left hand side of the
declaration.
Projection types do not appear in type annotations, but Typed Lua
uses them to project unions of tuple types into unions of
first-level types that have a dependency relation.
We will discuss projection types in more detail in the next chapter.

So far, Typed Lua replaces projection types with the union of the
corresponding component of each tuple, when it infers that a variable
has a projection type.
In our example, the variables \texttt{q} and \texttt{r} get projection
types that map to the first and second components of the union of
tuple types \texttt{(integer, integer, nil*)|(nil, string, nil*)}.
This means that variables \texttt{q} and \texttt{r} have types
\texttt{integer|nil} and \texttt{integer|string}, respectively.

If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates against all tuples in the projected union.
In our example, the projected union has type
\texttt{(integer, integer, nil*)|(nil, string, nil*)} outside of the
\texttt{if} statement,
but \texttt{(integer, integer, nil*)} inside the \texttt{then} block,
and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
Thus, variable \texttt{q} has type \texttt{integer|nil} and variable
\texttt{r} has type \texttt{integer|string} outside of the
\texttt{if} statement;
but variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
has type \texttt{integer} inside the \texttt{then} block;
and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
has type \texttt{string} inside the \texttt{else} block.
